UDC 911.3

### 2024-03-06 legal-companies-concentration-and-availability/Paper.Rmd
Proofreading (round 2) and other minor improvements

**Abstract:** the article examines the relationship between the concentration of legal companies and the availability of legal services. To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia. The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies. The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia. Availability is measured as the number of employees of legal companies per 100,000 population. Data visualization and linear regression were utilized to estimate the hypothesized relationship. The study showed a positive linear relationship between concentration and availability. However, there are multiple approaches to achieving good availability of legal services.

**Keywords:** legal services, concentration, availability, small and medium-sized business, open data

@@

Cities have traditionally been viewed as regional centers that provide goods and services to smaller surrounding settlements [@Christaller1966]. It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019]. Concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining. However, an increase in the number of firms does not necessarily lead to a corresponding growth in the availability of services they provide. The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996], because it is directly connected to the quality of life and human rights. Research has also been conducted on the links between market properties and service availability [@Lu2020]. Typically, the literature focuses on medical services [@Joseph1982], but other professional services are also worth attention, such as legal services [@Blacksell1988; @Patel2008].

This paper empirically examines link between the concentration of legal companies in Russian cities and the availability of legal services. The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by the Federal Tax Service (FTS). The main research question is how the availability of legal professionals is affected by the concentration of legal companies. Additionally, the study aims to identify any geographical patterns in the spatial distribution of cities based on legal services availability.

**Data.** In this research, a company-wise geo-referenced dataset on small and medium-sized legal companies in Russia was used. The dataset was created using the Python CLI tool *ru-smb-companies* developed by the author of this paper[^1]. The tool downloads, extracts, aggregates and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a compact CSV dataset. To select legal companies, a filtering by main activity code equal to 69.10 was applied. Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue for the target year, and the company's address of registration is in a city rather than in a rural area. During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of companies or employees per 100,000 population and were subsequently excluded from the data.

@@

**Methods.** The *concentration of legal companies* in cities was calculated by dividing the number of employees of legal companies in a given city by the total number of employees of legal companies in all cities. This formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021]. The *availability of legal services* was measured by dividing the number of employees of legal companies in a city divided by the city's population. This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. To categorize the cities by population, the classification system of Lappo [@Lappo1997] was used. The main method employed was data visualization, including mapping, with linear regression used as a supplementary method. The analysis was performed using the R programming language and additional packages. The source code for this paper can be found at the public repository[^2].

@@

**Results and discussion.** Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in Methods section. The plot displays the groups by city population and city rank within the region, in addition to the general trend. The overall connection between the concentration and availability appears to be linear. The linear model results demonstrate that the relationship between the two variables can be described by the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.

Examining the regression line common to all data points, we see that while the linearity is generally good and the linear modeling results are satisfactory, the total distribution is fan-shaped. This means that the variance of error of the linear model decreases as both the dependent and independent variables increase. However, this is mainly due to the presence of several distinct population size groups of cities in the data. Within each of these groups, linear models satisfy the assumptions better. The link between concentration and availability is similar across all city population groups, and the regression lines are nearly parallel, indicating equal slope coefficients and differing only in intercepts. Additionally, it can be concluded that, in general, larger cities have higher concentrations and availability, although the relationship is relatively weak.

@@

```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1. The relationship between the concentration of legal companies and the availability of legal services. City population group and relative city rank within the region are marked. The formulas used to calculate the metrics are described in the Methods section. Both the X and Y axes are log10-scaled. Dotted regression lines represent the linear fits within city population groups, while the solid gray line represent the overall linear fit", fig.height=4}
ca_plot
```

In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services. One possible reason for this is a considerably big size of such cities.

Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability. Moscow alone accounts for almost one-third of all legal companies employees, and Saint Petersburg accounts for about 10%. The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.

@@

Figure 2 displays the spatial distribution of cities based on the availability of legal services. Regional centers exhibit a high relative level of availability. The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities. Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg in the Urals, and Tuymen in Western Siberia. There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.

@@

**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the availability of legal services. This trend is observed in all the city population groups. The largest cities in regions typically have relatively high availability of legal services, as do cities with population above 1 million. Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country. Empirical data suggests that there may be a saturation limit beyond which the increase of concentration do not have a substantial impact on availability. There are multiple paths to achieving high levels of legal services availability, with increasing concentration of legal companies in a city being just one option. Even cities with relatively low concentration can achieve similar levels of availability as larger ones. Moscow and its region are a cluster of areas with high legal services availability. Similarly, Saint Petersburg, Yekaterinburg, Krasnodar, Tyumen, and Stavropol also have good access to legal services.

>>>>

**Abstract:** The article examines the relationship between the concentration of legal companies and the availability of legal services. To conduct the analysis, a dataset was created using open data from the Federal Tax Service of Russia. The dataset includes information on the location, revenue, expenditure, and number of employees of small and medium-sized legal companies. The concentration is calculated by dividing the number of employees of legal companies in the city by the total number of employees of all legal companies in Russia. Availability is measured as the number of employees of legal companies in the city per 100,000 population. Data visualization and linear regression were utilized to estimate the hypothesized relationship. The study showed a positive linear relationship between concentration and availability. However, there are multiple approaches to achieving good availability of legal services.

**Keywords:** legal services, concentration, availability, small and medium-sized businesses, open data

@@

Cities have traditionally been viewed as regional centers that provide goods and services to smaller surrounding settlements [@Christaller1966]. It is well-known that professional and business services tend to cluster in large cities, and their spatial distribution is closely linked to urban hierarchies [@Keeble2002; @Rubalcaba2013; @Kekezi2019]. The concentration of businesses provides economic benefits for the companies and thus is likely to become self-sustaining. However, an increase in the number of firms does not necessarily lead to a corresponding growth in the availability of services they provide. The issue of access to various services has been of great importance for decades [@Blacksell1990; @Keleher1996] because it is directly connected to the quality of life and human rights. Research has also been conducted on the links between market properties and service availability [@Lu2020]. Typically, the literature focuses on medical services [@Joseph1982], but other professional services are also worth attention, such as legal services [@Blacksell1988; @Patel2008].

This paper empirically examines the link between the concentration of legal companies in Russian cities and the availability of legal services. The analysis is based on information about the location of small and medium-sized legal businesses in Russia obtained from open data published by the Federal Tax Service (FTS). The main research question is how the availability of legal professionals is affected by the concentration of legal companies. Additionally, the study aims to identify any geographical patterns in the spatial distribution of cities based on legal services availability.

**Data.** In this research, a company-wise geo-referenced dataset on small and medium-sized legal companies in Russia was used. The dataset was created using the Python CLI tool *ru-smb-companies* developed by the author of this paper[^1]. The tool downloads, extracts, aggregates, and geocodes large archived XML dumps from the state registry of small and medium-sized businesses (SMBs) distributed by FTS, producing a compact CSV dataset. To select legal companies, a filtering by main activity code equal to 69.10 was applied. Additionally, the dataset was filtered by the following conditions: the year is 2021, the company is an organization rather than a sole entrepreneur, the company has declared non-zero revenue and expenditure for the target year, and the company's address of registration is in a city rather than in a rural area. During the exploratory data analysis, four cities (Innopolis, Kirovsk, Bronnytsy, Krasnoarmeisk) were identified as outliers based on the number of companies or employees per 100,000 population and were subsequently excluded from the data.

@@

**Methods.** The *concentration of legal companies* in cities was calculated by dividing the number of employees of legal companies in the given city by the total number of employees of legal companies in all cities. This formula was derived and adapted from the Atlas of Economic Specialization of Russian Regions [@Atlas2021]. The *availability of legal services* was measured by dividing the number of employees of legal companies in the city by the city's population. This formula is a modified version of the legal market saturation metric used in the analysis of the Russian legal services market conducted by the Institute for the Rule of Law [@Moiseeva2016]. To categorize the cities by population, the classification system of Lappo [@Lappo1997] was used. The main method employed was data visualization, including mapping, with linear regression used as a supplementary method. The analysis was performed using the R programming language and additional packages. The source code for this paper can be found in the public repository[^2].

@@

**Results and discussion.** Figure 1 shows the relationship between the concentration of legal companies and the availability of legal services, as measured by the metrics proposed in the Methods section. The plot displays the groups by city population and city rank within the region, in addition to the general trend. The overall connection between the concentration and availability appears to be linear. The linear model results demonstrate that the relationship between the two variables can be described by the formula *`r ca_model_formula`*, where both the intercept and slope are highly significant *(p < 0.001)*, and the explained variance is relatively high *(R^2^ = `r round(ca_model_summary$adj.r.squared, 2)`)*.

Examining the regression line common to all data points, we see that while the linearity is generally good and the linear modeling results are satisfactory, the total distribution is fan-shaped. This means that the variance of error of the linear model decreases as both the dependent and independent variables increase. However, this is mainly due to the presence of several distinct population-size groups of cities in the data. Within each of these groups, linear models satisfy the assumptions better. The link between concentration and availability is similar across all city population groups, and the regression lines are nearly parallel, indicating equal slope coefficients and differing only in intercepts. Additionally, it can be concluded that, in general, larger cities have higher concentrations and availability, although the relationship is relatively weak.

@@

```{r fig-1-concentration-and-accessibility, fig.cap="Figure 1. The relationship between the concentration of legal companies and the availability of legal services. City population group and relative city rank within the region are marked. The formulas used to calculate the metrics are described in the Methods section. Both the X and Y axes are log10-scaled. Dotted regression lines represent the linear fits within city population groups, while the solid gray line represents the overall linear fit. The regression line for the millionaire cities is fitted excluding the outliers (Moscow and Saint Petersburg)", fig.height=4}
ca_plot
```

In general, cities with the top rank by the population in their respective regions tend to have better availability of legal services. One possible reason for this is the considerably big size of such cities.

Moscow and Saint Petersburg are two remarkable outliers with extremely high concentration of legal services and a good level of legal services availability. Moscow alone accounts for almost one-third of all legal companies' employees, and Saint Petersburg accounts for about 10%. The position of the corresponding data points relative to the main regression lines suggests that there is an “availability saturation” level, where further concentration has little effect on availability.

@@

Figure 2 displays the spatial distribution of cities based on the availability of legal services. Regional centers exhibit a high relative level of availability. The most prominent cluster of good access to legal services is formed by Moscow and its satellite cities. Other notable centers with high availability include Krasnodar and Stavropol in the southern part of Russia's West, Yekaterinburg and Chelyabinsk in the Urals, and Tuymen and Novosibirsk in Western Siberia. There are regions with limited access to legal services, such as the bordering regions in the south of Siberia, as well as the Komi and Kalmyk Republics.

@@

**Conclusion.** The concentration of legal companies in cities has a positive linear relationship with the availability of legal services. This trend is observed in all the city population groups. The largest cities in regions typically have relatively high availability of legal services, as do cities with population above 1 million. Moscow and Saint Petersburg are characterized by the highest concentration and high availability, with Moscow alone accounting for almost one-third of all legal companies employees in the country. Empirical data suggests that there may be a saturation limit beyond which the increase in concentration does not have a substantial impact on availability. There are multiple paths to achieving high levels of legal services availability, with the increasing concentration of legal companies in a city being just one option. Even cities with relatively low concentration can achieve similar levels of availability as larger ones. Moscow and its region are a cluster of areas with high legal services availability. Similarly, Saint Petersburg, Krasnodar, Stavropol, Yekaterinburg, Chelyabinsk, Tyumen, and Novosibirsk also have good access to legal services.
